Parsing Postfix log files

### 2009-03-13 doc/logparser-implementation.tex
Corrections to flowchart section of implementation chapter. Improvements to graph size so they fill the width of the page. More consistent formatting of references to the flowchart. Alphabetical ordering for the actions with the same number of rules in the action distribution graph. Added more XXX items. Dealt with a few XXX items, particularly in the actions in PLP section of the implementation chapter. Improved the appearance of the flowchart a little. Updated the number of actions so that it is correct.

XXX WRITE AN INTRODUCTION\@.

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  XXX REFER TO
BLACK BOX PAPER\@?  There are some advantages to treating Postfix as a
black box during parser development:

@@

XXX MIGHT NEED A SMALL INTRO\@.

@@

    \item If the mail is rejected, save all data and finish.

    \item Follow the progress of the accepted mail until it is either
        delivered, bounced, or deleted, then save all data, and finish.

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: PICKUP\@; flowchart:~2).  Re-injection due to forwarding
sadly lacks explicit log lines of its own; it is explained in
\sectionref{Re-injected mails}.

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics as described in \sectionref{identifying
bounce notifications}.  Bounce notifications are the primary example of
internally generated mails, though there may be other types.  Postfix may
generate mails to the administrator when it encounters configuration
errors, but such mails are presumably rare.

\acronym{SMTP} is more complicated than local submission:

@@

    \item First there is a connection from the remote client (action:
        CONNECT\@; flowchart:~1).

    \item This is followed by rejection of sender address, recipient
        addresses, client \acronym{IP} address or hostname, HELO hostname,
        etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
        more mails (action: CLONE\@; flowchart:~5); or some interleaving of
        both.

    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
        Postfix has rejected any \acronym{SMTP} commands the data will be
        saved to the database; if not there will not be any data to save
        (any mails accepted will already have been cloned so their data is
        in another data structure).

@@

    \item Postfix will log the sender and recipient addresses separately
        (action: SAVE\_DATA\@; flowchart:~9).

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
        \sectionref{tracking re-injected mail} discusses this in
        detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: COMMIT\@; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from the state tables.
        Re-injected mails are described in \sectionref{tracking re-injected
        mail}.

@@

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator, \textit{unless\/} the mail is either parent
or child of re-injection, as explained in \sectionref{tracking re-injected
mail}.

@@

@@

XXX INTRODUCTION\@.

Actions are the component of this architecture responsible for processing
all of the inputs recognised by rules; in \parsername{} they reconstruct
the journey each mail takes through Postfix, dealing with all the
complications and difficulties that arise.  \parsername{} has
\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
distribution of rules to actions as shown in \graphref{Distribution of
rules per action}.  Unsurprisingly, the action with the most associated
rules is \texttt{DELIVERY\_REJECTED}, the action that processes Postfix
rejecting a mail delivery attempt; next is \texttt{SAVE\_DATA}, the action
responsible for processing informative log lines, supplementing the data
gathered from other log lines.  The third most common action is, perhaps
surprisingly, \texttt{UNINTERESTING}: this action does nothing when
invoked, allowing uninteresting log lines to be parsed without any effects
(it does not imply that the input is ungrammatical or unparsed).  Generally
rules specifying the \texttt{UNINTERESTING} action recognise log lines that
are not associated with a specific mail, e.g.\ notices about configuration
files changing; these log lines are recognised and processed so that the
framework can warn about unrecognised log lines, informing the user that
they need to add to or extend their ruleset.  The remaining actions have
only one or two associated rules: some are specialised actions required to
address a deficiency in the log files, or a complication that arises during
parsing; other actions will only ever have one log line variant, e.g.\ all
log lines showing that a remote client has connected are recognised by a
single rule and handled by the \texttt{CONNECT} action.  \parsernames{}
actions are described in detail in \sectionref{actions in detail in
implementation}, and \sectionref{adding new actions in implementation}
covers the process of adding a new action.

@@

    \item [MAIL\_BOUNCED] XXX WRITE THIS\@; INVOKES SAVE\_DATA ACTION\@.

    \item [MAIL\_DISCARDED] XXX WRITE THIS\@; INVOKES COMMIT ACTION\@.

@@

    \item [MAIL\_SENT] XXX WRITE THIS\@; INVOKES SAVE\_DATA ACTION\@.

>>>>

XXX WRITE AN INTRODUCTION\@.  MOST OF THIS IS CRAP, BUT I MIGHT BE ABLE TO
RESCUE SOME OF IT\@.

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Thus far
analysis of the log files has been sufficient to reconstruct Postfix's
behaviour, but for more difficult programs the techniques described in
\cite{black-box-error-reporting} may be useful.  There are some advantages
to treating Postfix as a black box during parser development:

@@

XXX NEEDS AN INTRO\@.

@@

    \item If the mail is rejected, save all data to the database and
        delete the connection entry from the state table.

    \item Follow the progress of the accepted mail until it is either
        delivered, bounced, or deleted, then save all data to the database,
        and delete the mail's entry from the state table.

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged \flowchart{PICKUP}{2}.  Re-injection due to forwarding sadly lacks
explicit log lines of its own; it is explained in \sectionref{Re-injected
mails}.  Internally generated mails lack any explicit origin in Postfix
2.2.x and must be detected using heuristics as described in
\sectionref{identifying bounce notifications}.  Bounce notifications are
the primary example of internally generated mails, though there may be
other types.  Postfix may generate mails to the administrator when it
encounters configuration errors, but such mails are presumably rare.

\acronym{SMTP} is more complicated than the others:

@@

    \item First there is a connection from the remote client
        \flowchart{CONNECT}{1}.

    \item This is followed by rejection of sender address, recipient
        addresses, client \acronym{IP} address or hostname, HELO hostname,
        etc.\ \flowchart{REJECTION}{4}; acceptance of one or more mails
        \flowchart{CLONE}{5}; or some interleaving of both.

    \item The client disconnects \flowchart{DISCONNECT}{6}.  If Postfix has
        rejected any \acronym{SMTP} commands the data will be saved to the
        database; if not there will not be any data to save (any mails
        accepted will already have been cloned so their data is in another
        data structure).

@@

    \item Postfix will log the sender and recipient addresses separately
        \flowchart{SAVE\_DATA}{9}.

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail \flowchart{TRACK}{10};
        \sectionref{tracking re-injected mail} discusses this in detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator \flowchart{COMMIT}{12}.  This is the last log line
        for this particular mail (though it may be indirectly referred to
        if it was re-injected).  If it is neither parent nor child of
        re-injection the data is cleaned up and entered in the database,
        then deleted from the state tables.  The handling of re-injected
        mails is described in \sectionref{tracking re-injected mail}.

@@

It should be emphasised that the mail delivery process above happens
whether the mail is delivered to a mailbox, piped to a command, delivered
to a remote server, bounced (due to a mail loop, delivery failure, or five
day timeout), or deleted by the administrator, \textit{unless\/} the mail
is either parent or child of re-injection, as explained in
\sectionref{tracking re-injected mail}.

@@

XXX GENERIC RULES\@.

XXX COMPRESSED FILES\@.

XXX STATE TABLES\@.

XXX LOAD AND VERIFY RULES\@.

@@

Actions are the component of this architecture responsible for processing
all of the inputs recognised by rules; in \parsername{} they reconstruct
the journey each mail takes through Postfix, dealing with all the
complications and difficulties that arise.  \parsername{} has
\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
distribution of rules to actions as shown in \graphref{Distribution of
rules per action}.  Unsurprisingly, the action with the most associated
rules is \texttt{DELIVERY\_REJECTED}, the action that processes Postfix
rejecting a mail delivery attempt.  The next most common action is, perhaps
surprisingly, \texttt{UNINTERESTING}: this action does nothing when
invoked, allowing uninteresting log lines to be parsed without any effects
(it does not imply that the input is ungrammatical or unparsed).  Generally
rules specifying the \texttt{UNINTERESTING} action recognise log lines that
are not associated with a specific mail, e.g.\ notices about configuration
files changing; these log lines are recognised and processed so that the
framework can warn about unrecognised log lines, informing the user that
they need to add to or extend their ruleset.  Most of the remaining actions
have only one or two associated rules: some are specialised actions
required to address a deficiency in the log files, or a complication that
arises during parsing; other actions will only ever have one log line
variant, e.g.\ all log lines showing that a remote client has connected are
recognised by a single rule and handled by the \texttt{CONNECT} action.
\parsernames{} actions are described in detail in \sectionref{actions in
detail in implementation}, and \sectionref{adding new actions in
implementation} covers the process of adding a new action.

@@

    \item [MAIL\_BOUNCED] This action behaves exactly like the
        \texttt{SAVE\_DATA} action; it saves all data captured by the
        recognising rule's regex, and does nothing more.  It is a separate
        action to help distinguish delivery attempts that bounce from other
        delivery attempts.

    \item [MAIL\_DISCARDED] Sometimes mail is discarded, e.g.\ mail
        submitted locally that is larger than the limit configured by the
        administrator.  This action is used for those mails; it invokes the
        \texttt{COMMIT} action, but is a separate action to simplify
        further analysis.

@@

    \item [MAIL\_SENT] This action behaves exactly like the
        \texttt{SAVE\_DATA} action; it saves all data captured by the
        recognising rule's regex, and does nothing more.  It is a separate
        action to help distinguish successful delivery attempts from other
        delivery attempts.
